%\section{References}

\begin{thebibliography}{99}

\bibitem{SR117}
Board of Governors of the Federal Reserve System (2011).
\textit{Supervisory Guidance on Model Risk Management (SR 11-7)}.
Federal Reserve Bulletin.

\bibitem{EIOPA2015}
European Insurance and Occupational Pensions Authority (2015).
\textit{Guidelines on the Use of Internal Models}.
EIOPA.

\bibitem{Sculley2015}
Sculley, D., Holt, G., Golovin, D., et al. (2015).
Hidden technical debt in machine learning systems.
In \textit{Advances in Neural Information Processing Systems}.

\bibitem{Molnar2022}
Molnar, C. (2022).
\textit{Interpretable Machine Learning} (3rd ed.).
Available at \url{https://christophm.github.io/interpretable-ml-book/}

\bibitem{Gama2014}
Gama, J., Žliobaitė, I., Bifet, A., Pechenizkiy, M., \& Bouchachia, A. (2014).
A survey on concept drift adaptation.
\textit{ACM Computing Surveys}, 46(4), 1--37.

\bibitem{Coifman2006}
Coifman, R. R., \& Lafon, S. (2006).
Diffusion maps.
\textit{Applied and Computational Harmonic Analysis}, 21(1), 5--30.

\bibitem{DeutschMedioni2015}
Deutsch, S., \& Medioni, G. (2015).
Intersecting manifolds: Detection, segmentation, and labeling.
In \textit{Proceedings of the International Joint Conference on Artificial Intelligence (IJCAI)}.

\bibitem{MumfordShah1989}
Mumford, D., \& Shah, J. (1989).
Optimal approximations by piecewise smooth functions and associated variational problems.
\textit{Communications on Pure and Applied Mathematics}, 42(5), 577--685.

\bibitem{Hamilton1989}
Hamilton, J. D. (1989).
A new approach to the economic analysis of nonstationary time series and the business cycle.
\textit{Econometrica}, 57(2), 357--384.

\bibitem{Luxburg2007}
von Luxburg, U. (2007).
A tutorial on spectral clustering.
\textit{Statistics and Computing}, 17(4), 395--416.

\bibitem{Verte2026}
Verte, X. (2026).
\textit{When to Segment: Topology-Aware Energy-Based Diagnostics for Structural Model Uncertainty}.
SSRN Working Paper.
Available at \url{https://doi.org/10.2139/ssrn.6081486}

\end{thebibliography}
